The Concurrent Calculi Formalisation Benchmark

### 2024-02-02 short-paper/short-paper.tex
Further reorganize paper.
I have added a comment before each section with a general outline of
the main points that I suggest could be in that section.

\begin{abstract}
  %% AM: implemented DZ
% when mechanising  message-passing concurrency using process calculi.

  POPLMark and POPLMark Re\-loaded sparked a flurry of
  work on machine-checked proofs, and fostered the adoption of proof
  mechanisation in programming language research. Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  For this reason, we propose a new collection of benchmark problems
  focused on the difficulties that typically arise when mechanising
  formal models of concurrent and distributed programming languages,
  such as process calculi.  Our benchmark problems address three key
  topics: linearity, scope extrusion, and coinductive reasoning.  The
  goal of this new challenge is to clarify, compare, and advance the
  state of the art, fostering the adoption of proof mechanisation in
  future research on message-passing concurrency.

@@

\section{Introduction}
The POPLMark challenge~\cite{POPLMark} accelerated progress in the development of proof assistants, libraries, and best practices by establishing a common ground for systematically evaluating mechanisation techniques.
In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs using logical relations.
These challenges were deliberately limited in scope, while their authors called for future investigations into such areas as coinduction and linear environments.
In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising message-passing concurrency using process calculi.

\alb{Is there a sharp difference among those papers?}
Several papers in this realm already include mechanised proofs (\eg~\cite{Castro-Perez2021, Castro2020, lmcs:9985, DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or discuss proof mechanisation techniques (\eg~\cite{Bengtson2016, Gay2001, Thiemann2019, DBLP:conf/forte/ZalakainD21}).
Our experience, however, is that choosing the appropriate proof mechanisation techniques and 
tools remains a significant challenge with a limited understanding of their 
respective trade-offs.  This often leads researchers toward a trial-and-error approach, 
resulting in sub-optimal and ad-hoc solutions. In turn, this increases the overall effort 
required for proof mechanisation, and hampers the reuse of techniques and results.

It is therefore our view that the research community in message-passing
concurrency (and process calculi, more specifically) will benefit from a standardised set
of challenge problems.
To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, and \emph{coinductive reasoning} for infinite behaviours.
These are certainly not the only factors that may cause issues, but they are fundamental to concurrency theory, and emerge in most mechanisations.

\todo{Rewrite this section to be more coherent.}

@@

The mechanisation of novel research results often requires addressing more than
just one of these aspects at once; we see the combination of techniques
as a next step beyond the scope of this challenge, as discussed further in \Cref{sec:going-beyond}.

Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects within concurrency theory?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

On the whole, this benchmark seeks to advance three main goals:

\begin{enumerate}[label=\textbf{(G\arabic*)},leftmargin=10mm]
\item\label{item:goal-comperison-accessibility} To enable the comparison of
  proof mechanisation approaches, making the challenges accessible to
  mechanisation experts who may be unfamiliar with concurrency theory;

@@

\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this section.}

In this section, we outline the factors considered in designing the
benchmark challenges. We begin with some general remarks, then
describe the individual design considerations for each challenge problem,
and the evaluation criteria.
% We have identified three key aspects fundamental to mechanisations of concurrent calculi and designed our challenge problems to exercise these.

@@

% While our challenges require reasoning about binders, which was the main topic of the original POPLMark, they also require other techniques.
% Our challenges also require reasoning about logical relations which goes above the reasoning involved in POPLMark Reloaded.

% Our challenge problems do not include reasoning about encodings. While
% this kind of reasoning is common in the field, it does not require any
% different reasoning techniques in a concurrency setting than in any
% other setting.

@@

\subsubsection{Coinduction.}
%\fxerror{
% ``Proper'' barbs with a structural congruence in coinduction instead of LTS semantics?
%}

Our coinduction challenge concerns strong barbed bisimulation and congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms,\alb{? why terms?} here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, % when considering the strong notions of equivalence,
thus simplifying the theory.
We have chosen not to include delegation in the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.

@@

\section{Going beyond the challenge problems}\label{sec:going-beyond}

Similar to the POPLMark and its Reloaded counterpart, this challenge
does not encompass all potential problems in the field. Our benchmark
is extendable in two aspects: it can further amalgamate and broaden
the existing challenges, and tackle aspects not yet covered by them.

% As with the POPLMark and POPLMark Reloaded challenges, this challenge does not aim to cover every possible
% problem in the area. Our benchmark can be extended in two
% dimensions: further expanding and combining the current challenges,
% and addressing features not covered by the current challenges.

\vspace{-1mm}%
\paragraph{Extending and combining the current challenges.}
While the proposed challenges try to minimise the overlap of the techniques
and aspects they explore (linearity, scope extrusion, coinduction), most
research on concurrency theory combines some or all of these. A second round of formalisations might
entail combining the features of the challenges.
We consider the following three extensions:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the
addition of channel delegation to those calculi that do not already
support it.%

\vspace{-1mm}%
\paragraph{Addressing new calculi and new features.}
Extending in this direction
means proposing new challenges covering different features of
process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in
further challenges could be multiparty session types and
choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. --
that we do not address here. Also, one could easily
design a challenge that goes beyond our coinduction challenge, by
exploring different notions of bisimilarity (barbed, weak, \etc)
and trace equivalence.
%
\luca{Coinduction may also play a relevant role in supporting recursive
processes and session types. Indeed, it is often the case that session type
systems (and logics) with recursive session types (and propositions) are
naturally expressed in \emph{infinitary form} by interpreting their typing (and
proof) rules coinductively. I CAN ADD ONE OR TWO REFERENCES IF YOU FEEL IT'S APPROPRIATE}

@@

Ultimately, the current challenges can be extended in several
worthwhile directions, and we look forward to a future when they
indeed are. We see the current challenge as setting
the foundation for those future extensions. It is our expectation that
the solutions of those future challenges will build on solutions
to this one.

>>>>

\begin{abstract}
  POPLMark and POPLMark Reloaded sparked a flurry of work on machine-checked proofs, and fostered the adoption of proof mechanisation in programming language research.
  Both challenges were purposely limited in scope, and their benchmark problems do not address concurrency-related issues.
  %
  We propose a new collection of benchmark problems focused on the difficulties that typically arise when mechanising formal models of concurrent and distributed programming languages, such as process calculi.
  Our benchmark problems address three key topics: linearity, scope extrusion, and coinductive reasoning.
  The goal of this new challenge is to clarify, compare, and advance the state of the art, fostering the adoption of proof mechanisation in future research on message-passing concurrency.

@@

% - Introduction
%   - POPLMark accelerated progress in mechanisation tools for PL
%     - POPLMark Reloaded tried to do the same for logical relations
%     - But they were deliberately limited in scope to not include process calculi
%   - We propose a new collection of benchmark problems for process calculi
%     - Why not concurrency in general? (after this point, only mention process calculi)
%     - Why is there a problem with process calculus mechanisations?
%     - Our experience says there is, and we will cite other experiences below
%   - We have noticed three key issues
%     - They are not covered by POPLMark
%     - There are fewer techniques than for binders
%     - Others also mention them as problems: cite and explain references briefly for each challenge topic
%   - Our goals are to determine the state of the art, develop best practices and tutorials, and identify improvements to proof assistant technology that would materially help the situation
\section{Introduction}
The POPLMark challenge~\cite{POPLMark} accelerated progress in the development of proof assistants, libraries, and best practices by establishing a common ground for systematically evaluating mechanisation techniques.
In doing so, it spearheaded a shift in the programming language research community towards publications that include mechanised proofs.
POPLMark Reloaded~\cite{POPLMarkReloaded} proposed a similar program for proofs using logical relations.
These challenges were deliberately limited in scope, while their authors called for future investigations into such areas as coinduction and linear environments.

In this spirit, we propose a new collection of benchmark problems specifically designed to address typical issues that arise when mechanising message-passing concurrency using process calculi. \todo{Why just process calculi and not concurrency in general?}
While many efforts to mechanise results about process calculi have been completed (as we will discuss further below), our experience is that choosing appropriate proof mechanisation techniques and tools remains a significant challenge with a limited understanding of the trade-offs.
This often leads researchers toward a trial-and-error approach, resulting in sub-optimal and ad-hoc solutions.
In turn, this increases the overall effort required for proof mechanisation, and hampers the reuse of techniques and results.

%\alb{Is there a sharp difference among those papers?}
%Several papers in this realm already include mechanised proofs (\eg~\cite{Castro-Perez2021, Castro2020, lmcs:9985, DBLP:journals/jar/CruzFilipeMP23, Tirore:2023}), or discuss proof mechanisation techniques (\eg~\cite{Bengtson2016, Gay2001, Thiemann2019, DBLP:conf/forte/ZalakainD21}).

It is therefore our view that the research community in message-passing concurrency (and more specifically, process calculi) will benefit from a standardised set of challenge problems.
To this end, our benchmark problems (discussed in detail in \cref{app:challenges}) are designed to focus on three key aspects which typically cause difficulties when mechanising concurrency theory: \emph{linearity}, \emph{scope extrusion}, and \emph{coinductive reasoning} for infinite behaviours.
The mechanisation of novel research results often requires addressing more than just one of these aspects at once; we see the combination of techniques as a next step beyond the scope of this challenge, as discussed further in \Cref{sec:going-beyond}.
These are certainly not the only factors that may cause issues, but they are fundamental to concurrency theory, and emerge in most mechanisations, as we will see presently while discussing the issues presented by the three key aspects.

@@

\subsection{Goals and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

To provide a framework in which to answer these questions, our benchmark is designed to satisfy three main goals:
\begin{enumerate}[label=\textbf{(G\arabic*)},leftmargin=10mm]
\item\label{item:goal-comperison-accessibility} To enable the comparison of
  proof mechanisation approaches, making the challenges accessible to
  mechanisation experts who may be unfamiliar with concurrency theory;

@@

% - Design of the benchmark
%   - We first list some general design remarks, then for each challenge
%   - General remarks
%     - We do not aim to be comprehensive (list other topics)
%     - However, we believe the best practices developed could also be of use in those topics
%     - We base our challenges on basic textbook theory with interesting proof techniques
%     - The challenges are independent (maybe note ``failure'' of POPLMark that everyone only did the first part)
%     - The challenges are as minimal as possible
%   - Linearity
%     - Type preservation for session types
%     - The main issue is context splitting
%     - Session types are pervasive in the field and simple to understand
%     - Binding two names makes duality obvious
%     - Brief overview of approaches and issues with them
%   - Scope extrusion
%     - Process calculi have several binders, but only restriction is different from ``normal'' ones
%     - The main issue is scope extrusion
%     - The pros and cons of established techniques for handling binders are different for process calculi, and it is unclear how exactly
%     - Brief overview of approaches and issues with them
%   - Coinduction
%     - Strong barbed bisimulation and congruence
%     - The main issue is the encoding of coinductive up-to techniques
%     - Strong equivalences to simplify (we don't have to abstract over internal transitions)
%     - Why barbed equivalences?
%     - No delegation to simplify
%     - Brief overview of approaches and issues with them
%   - Evaluation criteria
%     - We care not only about solutions, but about experiences and comparisons
%     - Mechanisation overhead
%     - Adequacy of the formal statements
%     - Cost of entry for tools and techniques
%     - Slight changes are OK if they are justified and adequate
\section{Design of the Benchmark}\label{sec:design-discussion}
\todo{Add references to support our claims and decisions in this section.}

In this section, we outline the factors considered in designing the benchmark challenges.
We begin with some general remarks, then describe the individual design considerations for each challenge problem, and the evaluation criteria for solutions.

@@

@@

\subsubsection{Coinduction.}
Our coinduction challenge concerns strong barbed bisimulation and congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms,\alb{? why terms?} here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, thus simplifying the theory.
We have chosen not to include delegation in the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.

@@

% - Going beyond the challenge problems
%   - We do not aim to be comprehensive, but the techniques should also be useful for other things
%   - Our challenges are independent, but they can also be combined
%   - Our challenges are basic textbook theory, so they can form the basis for many extensions
%   - Our aim is to see a future where ``the basics'' are as trivial in mechanisations as they are on paper
\section{Going beyond the challenge problems}\label{sec:going-beyond}
Similarly to POPLMark and its Reloaded counterpart, our challenges do not encompass all issues in the field.
Our benchmark is extendable in two aspects: one can further amalgamate and broaden the existing challenges, and tackle aspects not covered by the benchmark at all.
All of these extensions will still need to handle at least one of the key aspects of our challenges.

\paragraph{Extending and combining the current challenges.}
\todo{This paragraph is very long for how relatively little it says. Maybe remove/cut down?}
While we have tried to minimise the overlap of the techniques and aspects our challenges explore (linearity, scope extrusion, coinduction), most research on concurrency theory combines some or all of these.
A second round of formalisations might entail combining the features of the challenges.
We consider the following three extensions to be obvious places to start:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the addition of channel delegation to those calculi that do not already support it.%

\paragraph{Addressing new calculi and new features.}
\todo{Maybe cite some references in this section?}
Extending in this direction means proposing new challenges covering different features of process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in further challenges could be multiparty session types and choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. -- that we do not address here.
Also, one could easily design a challenge that goes beyond our coinduction challenge, by exploring different notions of bisimilarity (barbed, weak, \etc) and trace equivalence.
%
\luca{Coinduction may also play a relevant role in supporting recursive
processes and session types. Indeed, it is often the case that session type
systems (and logics) with recursive session types (and propositions) are
naturally expressed in \emph{infinitary form} by interpreting their typing (and
proof) rules coinductively. I CAN ADD ONE OR TWO REFERENCES IF YOU FEEL IT'S APPROPRIATE}

@@

Ultimately, the current challenges can be extended in several worthwhile directions, and we look forward to a future when they indeed are.
We see the current challenge as setting the foundation for those future extensions.
It is our hope and aim that our challenges will move us closer to a future where the key basic proof techniques are as easy to mechanise as they are to write on paper.
